---
title: "Senator Voting Records in the 113th Congress, First Session, 2013"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r libraries, echo=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
```




```{r data, echo=FALSE, include=FALSE}
senator_party <- read.csv(file="senator_party.csv")
senator_votes_1 <- read.csv(file="senator_votes_113-1.csv")
senator_votes_2 <- read.csv(file="senator_votes_113-2.csv")
senators <- read.csv(file="senator.csv")
result_code <- read.csv(file="result_code_list.csv")
tally_code <- read.csv(file="tally_code_list.csv")

regions_data <- read.csv(file="regions_data.csv")
regions_description <- read.csv(file="regions_description.csv")

regions_data <- subset(regions_data, id_regions_description != 5)
```

```{r combine table, echo=FALSE, include=FALSE}

vote_table <- select(senator_votes_1,voteno,member_code,tally_code) %>%
  filter(tally_code %in% c("A","B"))

vote_table$tally_code = ifelse(vote_table$tally_code=="A",0,1)

party_votes <- left_join(vote_table,select(senator_party,member_code,caucus_with), by = "member_code") 

#party_votes

party_vote_count <- party_votes %>%
  select(voteno, caucus_with, tally_code) %>%
  group_by(voteno, caucus_with) %>%
  mutate( total_party_votes = sum(voteno)/voteno ) %>%
  group_by(voteno, caucus_with, total_party_votes) %>%
  summarise( total_party_support = sum(tally_code) )

party_vote_count <- party_vote_count %>%
  mutate( avg_party_support = total_party_support/total_party_votes)

#party_vote_count
  

party_support_difference <- select(party_vote_count,voteno,caucus_with,avg_party_support) %>%
  spread(key = caucus_with, value = avg_party_support) %>%
  mutate( difference = D-R )

#party_support_difference



name_votes <- left_join(party_votes,select(senators,member_code,print_name,state_code), by = "member_code")

name_votes <- left_join(name_votes,select(regions_data,state_code,id_regions_description), by = "state_code")

name_votes <- left_join(name_votes,select(regions_description,id_regions_description,title), by = "id_regions_description")

#name_votes


group_votes <- select(name_votes,voteno,tally_code,caucus_with) %>%
  group_by(caucus_with) %>%
  summarise( total = sum(tally_code) )

#group_votes


```


## Average Support by Party

For each bill proposed in the senate, a senator votes Yea or Nea. By regarding Yae = 1 and Nae = 0 and restricting to senators in a given party, we may compute the average support of each party for a given bill by taking the mean of the votes of all senators in that party. The disparity between the parties can be visualized by plotting these average supports.


```{r plot party support, echo=FALSE}

party.colors = c(D='blue', R='red')

party_support_differenceD <- party_support_difference[order(party_support_difference$D),] %>%
  rowid_to_column()

party_support_differenceR <- party_support_difference[order(party_support_difference$R),] %>%
  rowid_to_column()

#party_support_differenceD

party_support_differenceD %>% ggplot() + geom_point( aes( x=as.factor(rowid), y=D, color='D')) + geom_point( aes( x=as.factor(rowid), y=R, color='R')) + xlab("Vote") + ylab("Party Support") + scale_color_manual(values=party.colors) + labs(color = "Party") + ggtitle("Average Support by Party, Increasing by Democratic Support") + theme(axis.text.x = element_blank())

party_support_differenceR %>% ggplot() + geom_point( aes( x=as.factor(rowid), y=D, color='D')) + geom_point( aes( x=as.factor(rowid), y=R, color='R')) + xlab("Vote") + ylab("Party Support") + scale_color_manual(values=party.colors) + labs(color = "Party") + ggtitle("Average Support by Party, Increasing by Republican Support") + theme(axis.text.x = element_blank())


party_support_difference %>% ggplot( ) + geom_point( aes(x=voteno, y=difference), color="green") + ggtitle("Difference in Average Support by Party") + xlab("Vote") + ylab("Republican Support          Democratic Support")

```
Notice that Democrats and Republicans usually vote differently.


## Average Supporty by Region

In addition to party, senators can be grouped into one of four regions based on their state: Northeast, South, Midwest, and West. As with the average party support, the average regional support of a bill can also be calculated by restricting to the senators in that region. Plotting shows the disparity in votes by region.


```{r plot regions support, echo=FALSE}

region.colors = c(Northeast='blue', South='red', Midwest='gold', West='green')

region_votes <- select(name_votes,voteno,print_name,tally_code,title)

region_vote_count <- region_votes %>%
  group_by(voteno, title) %>%
  mutate( total_region_votes = sum(voteno)/voteno ) %>%
  group_by(voteno, title, total_region_votes) %>%
  summarise( total_region_support = sum(tally_code) )

region_vote_count <- region_vote_count %>%
  mutate( avg_region_support = total_region_support/total_region_votes)

#region_vote_count


region_support_difference <- select(region_vote_count,voteno,title,avg_region_support) %>%
  spread(key = title, value = avg_region_support)

#region_support_difference


region_support_differenceNE <- region_support_difference[order(region_support_difference$Northeast),] %>%
  rowid_to_column()

region_support_differenceS <- region_support_difference[order(region_support_difference$South),] %>%
  rowid_to_column()

region_support_differenceW <- region_support_difference[order(region_support_difference$West),] %>%
  rowid_to_column()

region_support_differenceMW <- region_support_difference[order(region_support_difference$Midwest),] %>%
  rowid_to_column()



region_support_difference %>% ggplot() + geom_point( aes( x=voteno, y=Northeast, color='Northeast')) + geom_point( aes(x=voteno, y=South, color='South')) + geom_point( aes(x=voteno, y=Midwest, color='Midwest')) + geom_point( aes(x=voteno, y=West, color='West')) + xlab("Vote") + ylab("Region Support") + ggtitle("Average Support by Region") + scale_color_manual(values=region.colors) + labs(color = "Region")


region_support_differenceNE %>% ggplot() + geom_point( aes( x=as.factor(rowid), y=Northeast, color='Northeast')) + geom_point( aes(x=as.factor(rowid), y=South, color='South')) + geom_point( aes(as.factor(rowid), y=Midwest, color='Midwest')) + geom_point( aes(as.factor(rowid), y=West, color='West')) + xlab("Vote") + ylab("Region Support") + theme(axis.text.x = element_blank()) + ggtitle("Average Support by Region, Increasing by Northeastern Support") + scale_color_manual(values=region.colors) + labs(color = "Region")

region_support_differenceS %>% ggplot() + geom_point( aes( x=as.factor(rowid), y=Northeast, color='Northeast')) + geom_point( aes(x=as.factor(rowid), y=South, color='South')) + geom_point( aes(as.factor(rowid), y=Midwest, color='Midwest')) + geom_point( aes(as.factor(rowid), y=West, color='West')) + xlab("Vote") + ylab("Region Support") + theme(axis.text.x = element_blank()) + ggtitle("Average Support by Region, Increasing by Southern Support") + scale_color_manual(values=region.colors) + labs(color = "Region")

region_support_differenceMW %>% ggplot() + geom_point( aes( x=as.factor(rowid), y=Northeast, color='Northeast')) + geom_point( aes(x=as.factor(rowid), y=South, color='South')) + geom_point( aes(as.factor(rowid), y=Midwest, color='Midwest')) + geom_point( aes(as.factor(rowid), y=West, color='West')) + xlab("Vote") + ylab("Region Support") + theme(axis.text.x = element_blank()) + ggtitle("Average Support by Region, Increasing by Midwestern Support") + scale_color_manual(values=region.colors) + labs(color = "Region")

region_support_differenceW %>% ggplot() + geom_point( aes( x=as.factor(rowid), y=Northeast, color='Northeast')) + geom_point( aes(x=as.factor(rowid), y=South, color='South')) + geom_point( aes(as.factor(rowid), y=Midwest, color='Midwest')) + geom_point( aes(as.factor(rowid), y=West, color='West')) + xlab("Vote") + ylab("Region Support") + theme(axis.text.x = element_blank()) + ggtitle("Average Support by Region, Increasing by Western Support") + scale_color_manual(values=region.colors) + labs(color = "Region")



```

Regions with similar voting records tend to be clustered around each other. This behavior is especially apparent when ordering the data with respect to a given region. For example, in the graph ordered by Southern support, the Midwest and the South show similar support for the same bills, whereas the Northeast shows very differnt support.


```{r consitency, echo=FALSE, include=FALSE}

support_totals <- left_join(region_support_difference,select(party_support_difference,voteno,D,R), by = "voteno")

#support_totals

senator_consistency <- select(name_votes,member_code,print_name,caucus_with,state_code,title) %>%
  group_by(member_code,print_name,caucus_with,state_code,title) %>% summarise()

#senator_consistency


senator_consistency %>% mutate(consist_D = -1, consist_R = -1, consist_NE = -1, consist_S = -1, consist_MW = -1, consist_W = -1 )



for( name in senator_consistency$print_name ){
  
  name_index <- which(senator_consistency$print_name == name)
  
  local_senator_consistency <- filter(name_votes, print_name == name) %>%
  left_join(support_totals, by = "voteno")
  
  
  senator_consistency[name_index, "consist_D"] <- ( 1 - mean( abs( local_senator_consistency$D - local_senator_consistency$tally_code ) ) )
  
  senator_consistency[name_index, "consist_R"] <- ( 1 - mean( abs( local_senator_consistency$R - local_senator_consistency$tally_code ) ) )
  
  senator_consistency[name_index, "consist_NE"] <- ( 1 - mean( abs( local_senator_consistency$Northeast - local_senator_consistency$tally_code ) ) )
  
  senator_consistency[name_index, "consist_S"] <- ( 1 - mean( abs( local_senator_consistency$South - local_senator_consistency$tally_code ) ) )
  
  senator_consistency[name_index, "consist_MW"] <- ( 1 - mean( abs( local_senator_consistency$Midwest - local_senator_consistency$tally_code ) ) )
  
  senator_consistency[name_index, "consist_W"] <- ( 1 - mean( abs( local_senator_consistency$West - local_senator_consistency$tally_code ) ) )
  
  
}

#senator_consistency

senator_consistency <- mutate(senator_consistency, name_state = paste(print_name," (",caucus_with,"-",state_code,")", sep="") )

#senator_consistency


ShelbyD <- which(senator_consistency$print_name == "Shelby" & senator_consistency$caucus_with == "D" )

senator_consistency <- senator_consistency[-ShelbyD,]

colnames(senator_consistency)[colnames(senator_consistency)=="caucus_with"] <- "party"

 

#senator_consistency


```

## Individual Senators Compared by Party and Region

We are also interested in viewing how consistently an individual senator votes with a given party or region. By considering the difference between an individual senator's vote and the average vote, and taking a mean over all votes, we obtain a score between 0 and 1 for each senator with respect to each category. Organizing these results with a bar plot allows us to visualize how aligned each senator's voting record is with the average voting record of the party or region.


```{r bar charts, cache=TRUE, echo=FALSE, fig.width=15, fig.asp=0.7}

party.colors = c(D='blue', R='red')

senator_consistency %>% ggplot( mapping=aes( x=reorder(name_state, -consist_D), y=consist_D, fill=party) ) + geom_bar( stat="identity") + theme(axis.text.x = element_text(angle=90,hjust=1,vjust=1)) + scale_fill_manual(values=party.colors) + ggtitle("Alignment with D") + xlab("Senator") + ylab("Consistency with Average D Votes") + ylim(0,1)

senator_consistency %>% ggplot( mapping=aes( x=reorder(name_state, -consist_R), y=consist_R, fill=party) ) + geom_bar( stat="identity") + theme(axis.text.x = element_text(angle=90,hjust=1,vjust=1)) + scale_fill_manual(values=party.colors) + ggtitle("Alignment with R") + xlab("Senator") + ylab("Consistency with Average R Votes") + ylim(0,1)

senator_consistency %>% ggplot( mapping=aes( x=reorder(name_state, -consist_NE), y=consist_NE, fill=party) ) + geom_bar( stat="identity") + theme(axis.text.x = element_text(angle=90,hjust=1,vjust=1)) + scale_fill_manual(values=party.colors) + ggtitle("Alignment with Northeast") + xlab("Senator") + ylab("Consistency with Average NE Votes") + ylim(0,1)

senator_consistency %>% ggplot( mapping=aes( x=reorder(name_state, -consist_S), y=consist_S, fill=party) ) + geom_bar( stat="identity") + theme(axis.text.x = element_text(angle=90,hjust=1,vjust=1)) + scale_fill_manual(values=party.colors) + ggtitle("Alignment with South") + xlab("Senator") + ylab("Consistency with Average S Votes") + ylim(0,1)

senator_consistency %>% ggplot( mapping=aes( x=reorder(name_state, -consist_MW), y=consist_MW, fill=party) ) + geom_bar( stat="identity") + theme(axis.text.x = element_text(angle=90,hjust=1,vjust=1)) + scale_fill_manual(values=party.colors) + ggtitle("Alignment with Midwest") + xlab("Senator") + ylab("Consistency with Average MW Votes") + ylim(0,1)

senator_consistency %>% ggplot( mapping=aes( x=reorder(name_state, -consist_W), y=consist_W, fill=party) ) + geom_bar( stat="identity") + theme(axis.text.x = element_text(angle=90,hjust=1,vjust=1)) + scale_fill_manual(values=party.colors) + ggtitle("Alignment with West") + xlab("Senator") + ylab("Consistency with Average W Votes") + ylim(0,1)




```

Notice that the average Northeast and West votes are primarily aligned with the average Democratic votes; likewise, Southern votes are primarily aligned wtih Republican votes. The Midwest exhibits more interesting behavior since it includes many senators of both parties.

Also of interest is the fact that Senator Collins (R-MA) votes more consistently with the Democratic party than any other Republican senator, but she also represents a state in the Northeast, which is primarily Democratic.


## Regional Voting Consistency by Party

By comparing the average party support with the average regional support, we may view how aligned individual regions are with a given party. This is summarized by the bar chart below.


```{r party supporty by region, echo=FALSE}

party_region <- data.frame(matrix(, nrow=6, ncol=7))

colnames(party_region)[colnames(party_region)=="X1"] <- "category"
colnames(party_region)[colnames(party_region)=="X2"] <- "conD"
colnames(party_region)[colnames(party_region)=="X3"] <- "conR"
colnames(party_region)[colnames(party_region)=="X4"] <- "conNE"
colnames(party_region)[colnames(party_region)=="X5"] <- "conS"
colnames(party_region)[colnames(party_region)=="X6"] <- "conMW"
colnames(party_region)[colnames(party_region)=="X7"] <- "conW"

party_region[1, "category"] <- "D"
party_region[2, "category"] <- "R"
party_region[3, "category"] <- "Northeast"
party_region[4, "category"] <- "South"
party_region[5, "category"] <- "Midwest"
party_region[6, "category"] <- "West"

party_region[1, "conD"] <- ( 1 - mean( abs( support_totals$D - support_totals$D ) ) )
party_region[2, "conD"] <- ( 1 - mean( abs( support_totals$D - support_totals$R ) ) )
party_region[3, "conD"] <- ( 1 - mean( abs( support_totals$D - support_totals$Northeast ) ) )
party_region[4, "conD"] <- ( 1 - mean( abs( support_totals$D - support_totals$South ) ) )
party_region[5, "conD"] <- ( 1 - mean( abs( support_totals$D - support_totals$Midwest ) ) )
party_region[6, "conD"] <- ( 1 - mean( abs( support_totals$D - support_totals$West ) ) )

party_region[1, "conR"] <- ( 1 - mean( abs( support_totals$R - support_totals$D ) ) )
party_region[2, "conR"] <- ( 1 - mean( abs( support_totals$R - support_totals$R ) ) )
party_region[3, "conR"] <- ( 1 - mean( abs( support_totals$R - support_totals$Northeast ) ) )
party_region[4, "conR"] <- ( 1 - mean( abs( support_totals$R - support_totals$South ) ) )
party_region[5, "conR"] <- ( 1 - mean( abs( support_totals$R - support_totals$Midwest ) ) )
party_region[6, "conR"] <- ( 1 - mean( abs( support_totals$R - support_totals$West ) ) )

party_region[1, "conNE"] <- ( 1 - mean( abs( support_totals$Northeast - support_totals$D ) ) )
party_region[2, "conNE"] <- ( 1 - mean( abs( support_totals$Northeast - support_totals$R ) ) )
party_region[3, "conNE"] <- ( 1 - mean( abs( support_totals$Northeast - support_totals$Northeast ) ) )
party_region[4, "conNE"] <- ( 1 - mean( abs( support_totals$Northeast - support_totals$South ) ) )
party_region[5, "conNE"] <- ( 1 - mean( abs( support_totals$Northeast - support_totals$Midwest ) ) )
party_region[6, "conNE"] <- ( 1 - mean( abs( support_totals$Northeast - support_totals$West ) ) )

party_region[1, "conS"] <- ( 1 - mean( abs( support_totals$South - support_totals$D ) ) )
party_region[2, "conS"] <- ( 1 - mean( abs( support_totals$South - support_totals$R ) ) )
party_region[3, "conS"] <- ( 1 - mean( abs( support_totals$South - support_totals$Northeast ) ) )
party_region[4, "conS"] <- ( 1 - mean( abs( support_totals$South - support_totals$South ) ) )
party_region[5, "conS"] <- ( 1 - mean( abs( support_totals$South - support_totals$Midwest ) ) )
party_region[6, "conS"] <- ( 1 - mean( abs( support_totals$South - support_totals$West ) ) )

party_region[1, "conMW"] <- ( 1 - mean( abs( support_totals$Midwest - support_totals$D ) ) )
party_region[2, "conMW"] <- ( 1 - mean( abs( support_totals$Midwest - support_totals$R ) ) )
party_region[3, "conMW"] <- ( 1 - mean( abs( support_totals$Midwest - support_totals$Northeast ) ) )
party_region[4, "conMW"] <- ( 1 - mean( abs( support_totals$Midwest - support_totals$South ) ) )
party_region[5, "conMW"] <- ( 1 - mean( abs( support_totals$Midwest - support_totals$Midwest ) ) )
party_region[6, "conMW"] <- ( 1 - mean( abs( support_totals$Midwest - support_totals$West ) ) )

party_region[1, "conW"] <- ( 1 - mean( abs( support_totals$West - support_totals$D ) ) )
party_region[2, "conW"] <- ( 1 - mean( abs( support_totals$West - support_totals$R ) ) )
party_region[3, "conW"] <- ( 1 - mean( abs( support_totals$West - support_totals$Northeast ) ) )
party_region[4, "conW"] <- ( 1 - mean( abs( support_totals$West - support_totals$South ) ) )
party_region[5, "conW"] <- ( 1 - mean( abs( support_totals$West - support_totals$Midwest ) ) )
party_region[6, "conW"] <- ( 1 - mean( abs( support_totals$West - support_totals$West ) ) )


#party_region

#senator_consistency <- senator_consistency[-ShelbyD,]

region_no_party <- party_region[-1,]
region_no_party <- region_no_party[-1,]

#region_no_party



stacked_region <- gather(select(region_no_party,category,conD,conR), key="con_party", value="consistency", 2:3)

#stacked_region


stacked.colors=c(conD="blue",conR="red")

#region_no_party %>% ggplot() + geom_bar(mapping=aes( x=category, y=conD ), fill="blue", stat="identity", alpha=0.5, position="dodge") + geom_bar(mapping=aes( x=category, y=conR ), fill="red", stat="identity", alpha=0.5, position="dodge") + ggtitle("Senate Voting Consistency by Party and Region") + xlab("Region") + ylab("Average Voting Alignment") + ylim(0,1)


colnames(stacked_region)[colnames(stacked_region)=="con_party"] <- "party_consistency"

stacked_region %>%
  ggplot( mapping = aes( x=category, y=consistency, fill=party_consistency )) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.75)) + xlab("Region") + ylab("Averge Voting Alignment") + ggtitle("Senate Regional Voting Consistency by Party, 2013") + scale_fill_manual(values=stacked.colors) + ylim(0,1)



```


